# Clustering and classification

```{r echo = FALSE, message = FALSE}

setwd("~/GitHub/IODS-project")

library(tidyverse)
library(GGally)
library(MASS)
library(corrplot)
library(plotly)

boston <- as_tibble(MASS::Boston)

```

### Introduction to the data and data wrangling

In this chapter, we study Boston housing data from R package MASS, which contains records on housing values in suburbs of Boston. The data includes observations from 506 suburbs and has 14 variables. Variable definitions can be found in [here](https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html).

```{r}

str(boston)

```

Below is the summary of the variables in the data. There are 12 numeric and 2 integer variables in the data set and these all have different scale ranges. 


```{r}

summary(boston)

```

The instructions of this exercise suggested to present a graphical overview of the data and to describe the outputs and comment the distributions of the variables and their relationships. The graphical overview of the data gets very difficult to read because there are 14 variables in the data; plotting the variables pairwise leads to very small and hard to read output if functions introduced in the course are used (i.e. `GGAlly::ggpairs` or `pairs`). 

Therefore, a new function `corrplot::corrplot` is introduced. The correlation plot below shows that several of the 14 variables are positively or negatively correlated to varying degrees.

```{r}

cor_matrix <- cor(boston)

corrplot(cor_matrix, method = "circle", type = "upper", cl.pos = "b", tl.pos = "d", tl.cex = 0.6)

```

Because all the variables have different scales, let's scale the data so that all the variables have mean value of 0 and standard deviation of 1.

```{r}

boston_scaled <- as_tibble(scale(boston))

summary(boston_scaled)

```

Some data wrangling before moving on to linear discriminant analysis:

```{r}

# Creating a categorical crime variable

bins <- quantile(boston_scaled$crim)
crim_labels <- c("low", "med_low", "med_high", "high")

boston_scaled <- boston_scaled %>%
  mutate(crim = 
           cut(crim,
               breaks = bins,
               labels = crim_labels,
               include.lowest = TRUE))

# splitting the data to traning and test sets

n <- nrow(boston_scaled)
ind <- sample(n,  size = n * 0.8)

train <- boston_scaled[ind,]
test <- boston_scaled[-ind,]

Correct <- test$crim
test <- dplyr::select(test, -crim)

```


### Linear discriminant analysis (LDA)

Fitting and printing the LDA model.

```{r}

# fitting the model
lda.fit <- lda(crim ~ ., data = train)
lda.fit

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "grey", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crim)

# plot the lda results
plot(lda.fit, dimen = 2, col = classes, pch = classes, main ="LDA (bi)plot")
lda.arrows(lda.fit, myscale = 20)

```

Using test data set to predict the classes with the LDA model. 

The crosstable below shows how well the LDA model predicts the crime rate in Boston suburbs. The LDA model is very accurate in predicting suburbs with high crime. Additionally, concerning suburbs with actual crime rate from low to medium high, the LDA model predicts crime rate rather accurately in the same region even if the prediction is not as accurate as with suburbs with high crime rate. 

```{r}

# fitting the model 
lda.predict <- predict(lda.fit, newdata = test)

# predicted values
Predicted <- lda.predict$class

# table of predicted vs. correct classes
descr::crosstab(Correct, Predicted, prop.r = T, plot = F)

```

### K-means clustering

Below is the summary of euclidian distances between observations.

```{r}

# the data
boston <- as_tibble(MASS::Boston)
boston_scaled <- as_tibble(scale(boston))

# distance matrix 
dist_eu <- dist(boston_scaled, method = "euclidian")
summary(dist_eu)

```

Here is the first set of k-means clustering done with four clusters. For easier inspection, I only plotted the pairs of firts six variables in the data set.

```{r}

km <- kmeans(boston_scaled, centers = 4)

# plot the Boston dataset with clusters
pairs(boston_scaled[1:6], col = km$cluster)

```

I investigated the correct number of cluster by checking the total of within cluster sum of squares. After determining that the optimal number of clusters is two, I plotted the clusters again. 

Interpratetion: the clusters seem to vary in across several variables. Looking at the crime variable, which has been the center if interest in this exercise, it looks like that the other cluster (black), includes suburbs with higher crime rates. The other cluster, however, includes only low crime rate suburbs.

```{r}

# MASS, ggplot2 and Boston dataset are available
set.seed(123)

# determine the number of clusters
k_max <- 10

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(Boston, k)$tot.withinss})

# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line')

# k-means clustering
km <-kmeans(boston_scaled, centers = 2)

# plot the Boston dataset with clusters
pairs(boston_scaled[1:6], col = km$cluster)

```

### Super bonus

Preparing the data. 

```{r}

model_predictors <- dplyr::select(train, -crim)

# check the dimensions
dim(model_predictors)
dim(lda.fit$scaling)

```

The figure below plots the fitted LDA model in three dimensions. 

```{r}

# matrix multiplication
matrix_product <- as.matrix(model_predictors) %*% lda.fit$scaling
matrix_product <- as.data.frame(matrix_product)

plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type = 'scatter3d', mode = 'markers', colors = "Dark2")

```

Here is the same plot with the colors of the points representing crime rates of Boston suburbs. 

```{r}

plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type = 'scatter3d', mode = 'markers', color = train$crim, colors = "Dark2")

```

Lastly, I did the same plot but this time the points are colored by using the two categories found in the k-means clustering exercise. Comparing the two last plots visually shows that k-means clustering identified those suburbs (green dots) that had high or medium high crime rate in the previous plot. This works as a one way to validate the k-means clustering categories.

```{r}

km <-kmeans(boston_scaled[2:14], centers = 2)
clusters <- factor(km$cluster[ind])

plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type = 'scatter3d', mode = 'markers', color = clusters, colors = "Dark2")

```

